Efficiency

### 2018-09-09 Second.Rnw
more

That paper, and a subsequent blog post on the topic which made the paper publically available, attracted the interest of the Senate Economics Reference Committee Inquiry into the Future of Naval Shipbuilding in Australia.  Senator Chris Ketter asked written questions on notice to the Department of Defence requesting responses to the paper specifically, and related comments from public officials, with a return date of 3rd April 2018.  I was also invited to provide evidence to the committee in a public hearing on the 7th of June 2018.  In that public hearing, I stood by the conclusions of my substantive paper, which was: ``between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.

@@

\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{28 May 2012 - Senate Estimates}]
\textit{Senator Johnston directly challenges Rear Admiral Justin Moffit (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENMoffit}.}
\\
\\
\textbf{Senator JOHNSTON:} I am told that there are no pump jets on any conventional class submarines in the world.
\\
\textbf{Rear Adm. Moffitt:} That is quite right. It might be a very easy task for us to write it off. I do not know, because I have not done-
\\
\textbf{Senator JOHNSTON:} The French and the Russians have tried. No-one has used them. They are not being used on anything other than nuclear powered submarines-particularly the Americans.
\\
\textbf{Rear Adm. Moffitt:} If it does not look like it makes sense, I will rule it out. But I cannot give you the answer to that yet because we have not done the work.
\\
\textbf{Senator JOHNSTON:} Who had the idea to look at pump jets? Where did that come from? Did it come from capability group saying that we had better had a look at that?
\\
\textbf{Rear Adm. Moffitt:} I do not know the answer to that. I doubt that there is a clear answer.

@@

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{10 Feb 2016 - Sean Costello, CEO DCNS Australia}]
\textit{In a blog post\parencite{costello2016} for the ASPI Strategist prior to the announcement in April, Sean Costello describes the significance of the pump jet technology to their bid.}
\\
\\
``As a complete submarine power, we understand conventional propulsion, which is why we also understand that propulsion is but one part of the submarine puzzle. In designing to the Australian requirement, it should come as no surprise then that the conventionally powered Shortfin Barracuda Block 1A is only 5\% lighter than its nuclear cousin."
\\
\\
``There’s no better example of what I’m talking about than the offer from France to transfer to Australia sovereign control and use of pump-jet propulsion technology for the Shortfin Barracuda—technology resident only in France, the UK and the USA. Technology born from the French SSBN program a generation ago. The stealth and hydrodynamic performances of pump jet propulsion are of course classified and in Australia known only to DCNS and the Australian Government."

@@

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{8 Apr 2016 - Sean Costello, CEO DCNS Australia, and Gerard Autret, Chief Naval Architect of the Shortfin Barracuda}]
\textit{In a blog post\parencite{autret2016} for the ASPI Strategist prior to the announcement in April, Sean Costello describes a number of the major features and design approaches for the Shortfin Barracuda, including relating the choice of the pumpjet closely to an accoustic requirement to close with another submarine at higher speed.}
\\
\\
``The challenge for any attack submarine is to maintain a nearly silent acoustic signature at a speed necessary to manoeuvre within weapons range of the target. The nuclear Barracuda is designed to reduce radiated noise when operating at a speed sufficient in order to close a threat submarine undetected. Of particular importance is the pump-jet propulsor, which combines a rotor and stator within a duct to significantly reduce the level of radiated noise through the effects of wake harmonisation and avoidance of cavitation."
\end{tcolorbox}

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{May 2016 onwards - DCNS/Naval Group Website for Shortfin Barracuda}]
\textit{The official website for the submarine claims the advantages of the pumpjet in specific terms, drawing comparisons with "obsolete propeller technology" and "otherwise identical submarines".  Last accessed 4 Sept 2018 \parencite{naval2018}.}
\\
\\
``Pump jet propulsion means the Shortfin Barracuda can move more quietly than submarines with obsolete propeller technology. In a confrontation between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."

@@

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{4 October 2017 - Jean-Michel Billig, Executive Director Australian submarine Program, Naval Group}]
\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program speculates that convetional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.  Subsequent reports quote Naval Group shifting responsibility for the propulsion choice to the Australian Government.}
\\
\\
``The vessels may end up with conventional propellers as well as air independent propulsion, which helps to increase underwater endurance, Mr Billig said."
\\
\\
``However decisions on forms of propulsion are ultimately a matter for the Australian Government."\parencite{iggulden2017}

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{25 Oct 2017 - Hevre Guillou, CEO Naval Group}]
\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the global CEO of Naval group responds to a question from Brendan Nicholson asking whether the Australian submarine will be fitted with a pump jet, following Jean-Michel Billig's comments.}
\\
\\
``Naval Group is the manufacturer of both conventional propeller-driven submarines and larger nuclear submarines with the propulsor. The propulsor’s inclusion means the future submarine will move more quietly than submarines with obsolete propeller technology. In a contest between two otherwise identical submarines, the one with the pump jet always has the tactical advantage. The propulsor was an integral part of our bid. The Department of Defence says it intends to use the pump jet propulsion system."
\end{tcolorbox}

\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{31 Oct 2017 - Greg Sammut, head of SEA 1000}]
\textit{In an interview that was published in a post\parencite{sammut2017} on the ASPI Strategist the leader of the future submarine project for the RAN responds to a question from Brendan Nicholson asking whether there is any doubt about using the pump-jet.}
\\
\\
``We are using pump-jet propulsion that was offered by France. Some claims about the pump jet are misleading. It’s not purely for use in nuclear submarines. It can be efficient across the entire speed range, taking account of the submarine’s size, the speed required and stealth. It needs to be tuned to the hull of the submarine. The hull needs to be tuned to the pump jet."
\end{tcolorbox}

@@

\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{25 Oct 2017 - Senate Estimates}]
\textit{Senator Hanson directly challenges Rear Admiral Gregory Sammut (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENhanson}.}
\\
\\
\textbf{Rear Adm. Sammut:}   I think the point that we're trying to make here is that the proposal
offered by France included a pump jet.

\textbf{Rear Adm. Sammut:}   The proposal—

\textbf{Senator Hanson:}   This is why you've chosen them as the design—to go with the pump
jet. So it's not propellers, is it?

\textbf{Senator Hanson:}   And that's why you chose France's Naval Group to—

\textbf{Rear Adm. Sammut:}   No, that is not, alone, why we chose France....
\\
\\
\textit{Extensive remarks from Rear Adm. Sammut, and further questions from Senator Hanson about whether the proposal involves converting a nuclear submarine to a diesel sumbarine}
\\
\\
\textbf{Rear Adm. Sammut:}  ...  The effort is not to convert an
existing boat into a diesel-electric submarine; we're designing a submarine using many of the
learnings and references that come out of the Barracuda design. A good example there would
be something like hull diameter. If you're going to change the diameter of the hull of a
submarine, you're into a whole new set of calculations about the structural integrity and
strength of the submarine. If we can utilise the work that France has already done in having a
hull diameter that's very similar to Barracuda, we can leverage all of that design work, which
has not only gone into the Barracuda but also has been tested by France in their methods of
design and certification to get to this point.

@@

\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{15 Dec 2017 - Senate Estimates}]
\textit{Senator Rex Patrick asks Stephen Johnson (General Manager Submarines) about the pumpjet proposal, and produces documents suggesting they are less efficient at low speed.  The Defence Chief Scientist, Dr Zelinsky, and Chief of Defence Force, Air Chief Marshal Binskin, and Head of Capability and Sustainment Group, Kim Gillis are also involved in the discussion, defending the choice of the pumpjet.  Senator Patrick also specifically raises the concern about the impact of the weight of the pumpjet at the extremity of the submarine.   Full exchange in \ref{SENjohnson}.}
\\
\\
\textbf{Senator PATRICK:} Mr Johnson, there's been a lot of talk about pump jets. I've now tabled a document which has come from the United States, from Connecticut. You'd be very familiar with the organisation there. That basically describes the characteristics of pump jets, their pros and cons, as any piece of engineering equipment has. The document—and there's also another document that I've taken from an ASPI site—shows the efficiency characteristics of a pump jet. Both the US document—now declassified—and the document from the ASPI website suggest that, at low speed, pump jets are relatively inefficient compared to, say, a propeller. That may well be the reason why we see a pump jet on a Mark 48 torpedo but generally not on merchant ships and so forth. Would you care to comment on that?

>>>>

That paper, and a subsequent blog post \parencite{morrison2018a} on the topic which made the paper publically available, attracted the interest of the Senate Economics Reference Committee Inquiry into the Future of Naval Shipbuilding in Australia.  Senator Chris Ketter asked written questions on notice to the Department of Defence requesting responses to the paper specifically, and related comments from public officials, with a return date of 3rd April 2018.  I was also invited to provide evidence to the committee in a public hearing on the 7th of June 2018.  In that public hearing, I stood by the conclusions of my substantive paper, which was: ``between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.

@@

\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{28 May 2012 - Senate Estimates}]
\textit{Senator Johnston directly challenges Rear Admiral Justin Moffit (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENMoffit}. Emphasis added.}
\\
\\
\textbf{Senator JOHNSTON:} I am told that there are no pump jets on any conventional class submarines in the world.
\\
\textbf{Rear Adm. Moffitt:} That is quite right. It might be a very easy task for us to write it off. I do not know, because I have not done-
\\
\textbf{Senator JOHNSTON:} The French and the Russians have tried. No-one has used them. They are not being used on anything other than nuclear powered submarines-particularly the Americans.
\\
\textbf{Rear Adm. Moffitt:} \textbf{If it does not look like it makes sense, I will rule it out. But I cannot give you the answer to that yet because we have not done the work.}
\\
\textbf{Senator JOHNSTON:} Who had the idea to look at pump jets? Where did that come from? Did it come from capability group saying that we had better had a look at that?
\\
\textbf{Rear Adm. Moffitt:} I do not know the answer to that. I doubt that there is a clear answer.

@@

This encounter shows that the viability of pumpjets was challenged some time ago.  It is noteworthy at the time that RADM Moffit makes no attempt to defend the consideration, acknowledging even that it might be easy for them to "write it off".  It is also noteworthy the tone of confidence that Senator Johnston's challenge has, driving directly for an explanation of who had the original idea, which RADM Moffit was unable or unwilling to answer.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{10 Feb 2016 - Sean Costello, CEO DCNS Australia}]
\textit{In a blog post\parencite{costello2016} for the ASPI Strategist prior to the announcement in April, Sean Costello describes the significance of the pump jet technology to their bid. Emphasis added.}
\\
\\
``As a complete submarine power, we understand conventional propulsion, which is why we also understand that propulsion is but one part of the submarine puzzle. In designing to the Australian requirement, it should come as no surprise then that \textbf{the conventionally powered Shortfin Barracuda Block 1A is only 5\% lighter than its nuclear cousin.}"
\\
\\
``There's no better example of what I'm talking about than the \textbf{offer from France to transfer to Australia sovereign control and use of pump-jet propulsion technology for the Shortfin Barracuda}-technology resident only in France, the UK and the USA. Technology born from the French SSBN program a generation ago. The stealth and hydrodynamic performances of pump jet propulsion are of course classified and in Australia known only to DCNS and the Australian Government."

@@

The significance of the pumpjet in Frances bid was clearly acknowledged well before the public announcement of DCNS's success.  It also provides a clear indication that the size of the submarine already appeared to be quite well-known, and was only intended to be a very minor modification from the nuclear Barracuda.  For such a small variation in displacement, it might be very reasonably expected that the length of the submarine would vary (the easiest modification), but the diameter would not need to.  These comments also display quite clearly that the pumpjet technology in France's possession is recognised as common with technology in the US and the UK.  This rules out the likelihood that some special or unique pumpjet technology, better suited just to diesel-electric submarines, had been pioneered by France alone.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{8 Apr 2016 - Sean Costello, CEO DCNS Australia, and Gerard Autret, Chief Naval Architect of the Shortfin Barracuda}]
\textit{In a blog post\parencite{autret2016} for the ASPI Strategist prior to the announcement in April, Sean Costello describes a number of the major features and design approaches for the Shortfin Barracuda, including relating the choice of the pumpjet closely to an accoustic requirement to close with another submarine at higher speed.}
\\
\\
``The challenge for any attack submarine is to maintain a nearly silent acoustic signature at a speed necessary to manoeuvre within weapons range of the target. The nuclear Barracuda is designed to \textbf{reduce radiated noise when operating at a speed sufficient in order to close a threat submarine undetected. Of particular importance is the pump-jet propulsor}, which combines a rotor and stator within a duct to significantly reduce the level of radiated noise through the effects of wake harmonisation and avoidance of cavitation."
\end{tcolorbox}

This comment provides clear context for the importance of the pump-jet.  Consistent with the physics outlined in Reference \parencite{morrison2018} the acoustic advantage is only emphasised by DCNS as being significant at higher speeds.  Whether it is viable for a diesel-electric submarine (as opposed to the nuclear one mentioned here) to maintain such speeds to close with many types of targets is left unclear.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{May 2016 onwards - DCNS/Naval Group Website for Shortfin Barracuda}]
\textit{The official website for the submarine claims the advantages of the pumpjet in specific terms, drawing comparisons with "obsolete propeller technology" and "otherwise identical submarines".  Last accessed 4 Sept 2018 \parencite{naval2018}. Emphasis added.}
\\
\\
``Pump jet propulsion means the Shortfin Barracuda can move more quietly than submarines with obsolete propeller technology. In a confrontation between two \textbf{otherwise identical submarines}, the one with pump jet propulsion always has the \textbf{tactical} advantage."

@@

The official explanation of the designer and builders website makes a carefully worded explanation of the superiority of the pumpjet, by limiting the comparison to a "tactical" encounter with an "otherwise identical" submarine.  As discussed in \ref{SENmorrison}, this is the only \textit{possible} target that an elevated speed such as could be achieved by a diesel-electric submarine might assist in closing with.  The submarine would be at a substantial "strategic" disadvantage in such an encounter, since its lower dived endurance, dived range, and poorer indiscretion ratio dramatically increases the probability that the pumpjet propelled submarine would wind up being the hunted, rather than the hunter.  More significantly, the speeds achieved would not allow the submarine to close with any other target from behind such as almost all surface ships, or nuclear submarines.  This statement exhibits extreme care in the wording in order to be plausibly defensible, and is aguably misleading.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{4 October 2017 - Jean-Michel Billig, Executive Director Australian submarine Program, Naval Group}]
\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program acknowledges that convetional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.  Subsequent reports quote Naval Group shifting responsibility for the propulsion choice to the Australian Government.}
\\
\\
(Interview transcribed from Reference \parencite{lane2017}.)
Journalist: Does the pumpjet technology that is on Suffren remain viable for the Australian submarine?
Jean-Michel Billig: It could be viable...
Journalist: So it's not mandated, its...
Jean-Michel Billig: What is driving the choice of technology is basically the operational requirement...
Journalist: So we could end up with conventional propeller?
Jean-Michel Billig: Yes, yes, we could, we could, we could also end up with something different
\\
\\
``However decisions on forms of propulsion are ultimately a matter for the Australian Government."\parencite{iggulden2017}
\end{tcolorbox}

This crucial concession reveals a very different level of confidence in the pumpjet being suitable for the Australian diesel-electric submarine.  It also gives an important glimpse into the processes that both selected the winning bidder, as well as the processes for the selection of key technology components.  By revealing that the pumpjet has not yet been firmly decided upon, and that the technology choice is a product of the operational requirement, it makes it clear that the operational requirement has not yet been determined with sufficient precision to make the suitable technology choice clear to a ship-builder with ample experience in both technologies.  Moreover, it also reveals that this was also the case at the conlcusion of the CEP, and remained so well over a year after.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{25 Oct 2017 - Hevre Guillou, CEO Naval Group}]
\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the global CEO of Naval group responds to a question from Brendan Nicholson asking whether the Australian submarine will be fitted with a pump jet, following Jean-Michel Billig's comments. Emphasis added.}
\\
\\
``Naval Group is the manufacturer of both conventional propeller-driven submarines and larger nuclear submarines with the propulsor. The propulsor's inclusion means the future submarine will move more quietly than submarines with obsolete propeller technology. \textbf{In a contest between two otherwise identical submarines, the one with the pump jet always has the tactical advantage. The propulsor was an integral part of our bid.} The Department of Defence says it intends to use the pump jet propulsion system."
\end{tcolorbox}

The verbatim quotation of the carefully worded comparison between ``otherwise identical" submarines by the global CEO more than a year after they first appeared on the Australian DCNS website demonstrate a high degree of preparation, and continued careful word selection on the topic of the pumpjet.  This statement also further reveals nuances about the CEP selection process, simultaneously declaring that the propulsor (i.e. pumpjet) was ``integral" to their bid, whilst also deferring to what the Department of Defence ``intends" do use. This exposes a clear tension between how 'integral' the pumpjet was to the bid, and Defence's freedom to choose otherwise.  It also exposes a dramatic divergence between the level of emphasis that is given to the pumpjet as an important key feature of the proposal in public promotion (as highlighted in \cite{lane2017}) and a Defence decision which seemed to choose only to have the option of including the pumpjet, with the real decision to be deferreed.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{25 Oct 2017 - Senate Estimates}]
\textit{Senator Hanson directly challenges Rear Admiral Gregory Sammut (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENhanson}.}
\\
\\
\textbf{Senator Hanson:}   This is why you've chosen them as the design-to go with the pump
jet. So it's not propellers, is it?

\textbf{Rear Adm. Sammut:}   The proposal offered by France included a pump jet, and that's the
proposal that we've accepted. \textbf{A pump jet remains on offer, and that's what we intend to use
for the Future Submarine.}

\textbf{Senator Hanson:}   And that's why you chose France's Naval Group to-

\textbf{Rear Adm. Sammut:}   No, that is not, alone, why we chose France. There are a number of
things. I am on record as providing testimony at the Senate of the number of things we
considered in the selection of Naval Group as our international partner. There were a range of
factors, and they were all over the capability of the international partner to deliver a regionally
superior submarine and one which Australia would have the sovereign capacity to operate and
sustain throughout its life. There were a number of criteria in mind.
We looked at the capabilities. That includes the platform capabilities. We looked at the
ability of the international partners to work with the combat system integrator of our choice.
Those things in particular pertain to the capability of the boat. We looked at rough order of
magnitude costs and schedules to ensure the participants understood the scale of the program
that we were embarking upon and could give us insights into the manner in which they would
manage costs and schedules throughout the program. We looked at matters concerning
program management. \textbf{We looked at the design and the way that safety is incorporated into
the design. It's important that that was understood from the outset because, whilst the
international partner would be designing the submarine, it would be the Commonwealth that
would be responsible for certifying the submarine as safe and fit for operations, and we
needed to make sure that the certifying process would match the design philosophy.}
We looked at methods of sustainment. \textbf{We looked at the manner in which each of the
participants would support the establishment of a sovereign sustainment capability in
Australia.} Of course, that was a key lesson from the Collins program. Even though we built
the submarines in Australia, it took us some time to establish the sovereign capacity to sustain
those boats. Indeed, having done that now is the reason why we are able to deliver beyond
international benchmark availability. \textbf{We also looked into the way in which data would be
transferred to Australia, the manner in which intellectual property would also be managed in
the program.} We looked at the ability of the participants to manage risks. Again, throughout
this complex program, that would be a very important factor. I would say that we didn't just
pick a submarine with a pump jet; we picked an international partner that could meet a range
of requirements to achieve not only a regionally superior submarine but the sovereign
capacity to operate and sustain that submarine and to deliver that within the time frame that
would be necessary.
\\
\\
...
\\
\\
\textbf{Rear Adm. Sammut:}  ...  The effort is not to convert an
existing boat into a diesel-electric submarine; we're designing a submarine using many of the
learnings and references that come out of the Barracuda design. \textbf{A good example there would
be something like hull diameter. If you're going to change the diameter of the hull of a
submarine, you're into a whole new set of calculations about the structural integrity and
strength of the submarine.} If we can utilise the work that France has already done in having a
hull diameter that's very similar to Barracuda, \textbf{we can leverage all of that design work, which
has not only gone into the Barracuda but also has been tested by France in their methods of
design and certification to get to this point.}

\end{tcolorbox}
This dialogue in the Senate Estimates received media coverage highlighting Senator Hanson's apparent lack of understanding about the difference between a submarine's power source (nuclear or diesel) and propulsion (propeller or pumpjet).  However, she succeeded in applying sustained pressure to the question of the pumpjet, and the significance that it had in the selection of DCNS as a design partner.  This elicited a committment to the pumpjet that appeared hesitant in its wording, which noted that the pump jet ``remains on offer", which is what ``we intend to use".

RADM goes to length to explain that the pumpjet was not ``alone" the reason for selecting DCNS as a partner, and in doing so reveals important insights into Defence's thinking and process regarding the decision.  Crucially, he reveals that the 'diameter of the hull' is an important aspect of France's design reference in the Barracuda, pointing out that changing the diameter of the hull incurrs a ``whole new set of calculations about the structural integrity and strength of the hull."  Together with remarks about the ``way that safety is incorporated into the design" and concerns about the alignment of the certifying process with the ``design philosophy" it would appear that the hull diameter actually emerges as plausibly one of the most significant distinguishing features in the French bid, perhaps alongside the pumpjet.

This is alarming for two reasons.  The first is that it contradicts subsequent explanations regarding pumpjet efficiency being incorrectly assessed by outsiders because they failed to account for how how the hull would be 'tuned' to the pumpjet and the pumpjet 'tuned' to the hull.  If the hull diameter is already fixed and known as being identical to that of the nuclear Barracuda, then no adjustment ought to be necessary to a good working pumpjet design that is used to propel this platform on the hull's account.  The performance of the pumpjet on the relevant hull would be well understood by the manufacturer offering it, and presumeably available to the Australian government who were looking to acquire it.

The second is that is reveals a stunning naivety in terms of basic understanding of the heirachy of engineering challenges.  It may be true that there are very many complex and challenging engineering calculations involved in building a submarine. However, it is certainly not true that scaling up a pressure hull to a new diameter to meet the same fixed static load is amongst the most difficult, risky or complex.  Whilst extending the length of a submarine is undoubtedly even easier, adjusting the diameter still to meet a structural integrity requirement to meet a known static pressure (from depth) is a trivially simple calculation in comparison to the complexities of, for instance, re-designing the duct intake of blade arrangments of a pumpjet to maximise efficiency at a different speed whilst simultaneously attempting to avoid cavitation.  It was quite simply and successfully accomplished when the Collins Class submarine was developed by scaling up both the length and diameter of it's smaller predecessor.

That this ranks as a significant consideration in the mind of the chief of this program should be the cause of considerable alarm to those taking an interest the project.  Were it known to the competitors that their ability to achieve this trivial engineering requirement was in question, they would no-doubt be highly offended.  Any contender that was not considered competent to design a pressure hull of a new diameter should never have been invited to participate.  RADM Sammut should be called upon to clarify what he meant by ``we needed to make sure that the certifying process would match the design philosophy" because if indeed he meant that Defence had already decided that their own certifying process already precluded scaled-up diameter hulls, this represents a very bold departure from what might be considered normal risk assessment processes for major engineering projects, including those used in designing and certifying the Collins.  Furthermore, if the advantage was considered so substantial, he should be called upon to substantiate how significant increased difficulty in certification would have been, with respect to cost and schedule.

Amongst RADM Sammut's additional claims regarding the reason for choosing DCNS/Naval Group, are futher claims regarding the establishment of 'sovereign capability' and 'data transfer' to Australia.  This has significant bearing on the way that subsequent comments made by Defence and DCNS regarding the state of their understanding of pumpjets, and the method for their development for the Australian submarine.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{31 Oct 2017 - Greg Sammut, head of SEA 1000}]
\textit{In an interview that was published in a post\parencite{sammut2017} on the ASPI Strategist the leader of the future submarine project for the RAN responds to a question from Brendan Nicholson asking whether there is any doubt about using the pump-jet.}
\\
\\
``We are using pump-jet propulsion that was offered by France. Some claims about the pump jet are misleading. It's not purely for use in nuclear submarines. It can be efficient across the entire speed range, taking account of the submarine's size, the speed required and stealth. It needs to be tuned to the hull of the submarine. The hull needs to be tuned to the pump jet. We did not specify that the boat should have or not have air-independent propulsion or lithium-ion batteries, rather that it met our requirements."
\end{tcolorbox}

This comment confirms that the official policy is still to use the pumpjet, implying that the decision has indeed been made already.  This is at odds with the statement made only weeks earlier by Jean-Michel Billig.  This is also the first noteable reference to the inter-relationship between the hull and the pumpjet, implying that the pumpjet will indeed be modified in order to suit the hull, and that the hull will be modified to suit the pumpjet.  It also implies that the same pumpjet as France is fitting to the Suffren (i.e. nuclear barracuda) will not simply be inlcuded 'as is' on the Australian non-nuclear 'Shortfin' version.  Given that the length and diameter of the hull were announced as being already decided, and seemingly very similar to the nuclear version, it raises the very serious question as to whether/why the changed shape of the hull would necessitate modification of the complex pumpjet.  As pointed out in the authors testimony to the Senate \ref{SENmorrison}, the emphasis of this relationship, above the vastly different operational speed profile of the nuclear and non-nuclear submarines which would affect the propulsion system profoundly, raises serious questions about the origin of the comment.

@@

This corroborates that the comments made by RADM Sammut regarding efficiency were also made by the Department, under due consideration, to a Question on Notice.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{15 Dec 2017 - Senate Estimates}]
\textit{Senator Rex Patrick asks Stephen Johnson (General Manager Submarines) about the pumpjet proposal, and produces documents suggesting they are less efficient at low speed.  The Defence Chief Scientist, Dr Zelinsky, and Chief of Defence Force, Air Chief Marshal Binskin, and Head of Capability and Sustainment Group, Kim Gillis are also involved in the discussion, defending the choice of the pumpjet.  Senator Patrick also specifically raises the concern about the impact of the weight of the pumpjet at the extremity of the submarine.   Full exchange in \ref{SENjohnson}.}
\\
\\
\textbf{Senator PATRICK:} Mr Johnson, there's been a lot of talk about pump jets. I've now tabled a document which has come from the United States, from Connecticut. You'd be very familiar with the organisation there. That basically describes the characteristics of pump jets, their pros and cons, as any piece of engineering equipment has. The document-and there's also another document that I've taken from an ASPI site-shows the efficiency characteristics of a pump jet. Both the US document-now declassified-and the document from the ASPI website suggest that, at low speed, pump jets are relatively inefficient compared to, say, a propeller. That may well be the reason why we see a pump jet on a Mark 48 torpedo but generally not on merchant ships and so forth. Would you care to comment on that?

\textbf{Mr Johnson:} As we said just a little earlier, the design process-whether it is a propeller or propulsion-is very much keyed to the criteria that you look for. The first thing I said was speed. \textbf{No matter what kind of propeller or propulsion pump jet you design, it has to be optimised for a particular speed.} There are classes of surface ships that look to optimise cross-ocean transit. Physics usually keeps you around 17 to 18 knots, and then there is the cost of maintenance and the cost of repair and the fact that there's no particular stealth requirement. Typically, you will find that those ships will use an open-tip propeller. \textbf{However, other ships, cruise ships in particular, use an electric shrouded pod, which is a variation of a pump-jet technology.} Of course, the technology gets changed over time, so the older a document is, the more it's dated to that point in time.
\\
\\
...
\\
\\
\textbf{Mr Johnson:} We were talking about optimising for the speed and the characteristics of the pump jet. The decision-making process for submarines is a little more complicated because there's an acoustic performance aspect as well as the pure efficiency of the blade-propulsor-pump jet arrangement. \textbf{Without going into details, it's a decision that we make with Navy, but I'm confident, having done this a few times already, that we are going to find this pump-jet approach to be the optimum for our particular circumstances for this particular submarine.}
\\
\\
...
\\
\\
\textbf{Senator PATRICK:} Okay. I'll go back to you, Mr Johnson. Noting that you're saying that this is based on something that hasn't been to sea yet - it will go to sea on the Barracuda; presumably that's the submarine you're referring to-there's still a risk that's sitting there. As the guardian of the taxpayer, I know where you sit. I have a responsibility from the taxpayers' perspective to ask: do you have a fallback position if the data that comes from the French trials, or other trials that you might do, presents you with data that's inconsistent with the operational profile that you intend for the submarine?

\textbf{Mr Johnson:} Yes. \textbf{We will understand the performance data early enough in the design that we can accommodate whatever is required. And then, just to reassure you, it's really going to be an adjustment to the design of the pump jet, not a decision to leave it.} Without going into too much detail, the curve that you're referring to is about the efficiency of the propeller, not about its acoustic performance. As you know, there's nothing free in physics, so those are different-shaped curves. What we seek to do, as Chief of the Defence Force said, is to make the right balance of multiple dimensions of this now.

@@

This testimony contradicts the assurance provided in an October interview with RADM Sammut.  It is clear from multiple parts of Mr Johnson's statement that the decision on whether to use the pumpjet or not, has not yet been taken, since he is ``confident... that we are going to find this pump-jet approach to be the optimum...".  Whilst acknowledging that he doesn't yet have the performance data from the nuclear Barracuda class (which is not yet in the water), he gives assurance that it will be obtained early enough to ``accommodate whateve is required", which he gives assurance is only going to be adjustments to the design of the pumpjet.  This suggests that it is still not even known whether the pump-jet that was an 'integral' part of the bid was offered in an unmodified state, or whether it was known by France that it would need to be modified from existing designs, or whether what was offered was actually only a skill-set and capability for the design of a new pumpjet.

In responding to the evidence produced by Senator Patrick that pumpjets may have an efficiency issue at low speeds, Mr Johnson refers to an 'electric shrouded pod' which is used on cruise ships, which he claims is a modification of a pumpjet.  Nothing about this section of his testimony seems to have any reference of connection to reality.  Cross-ocean transit regularly occur at higher speeds than 17-18kt, as do cruise ships, which he mentions. These speeds are essentially unreacheable (or only extremely briefly sustainable) by a diesel electric submarine.  I can find no evidence of any cruise ships having any kind of shround around the tips of their propellers, or bearing any resemblance or relationship to a pumpjet.  As I pointed out in Mr Johnson's presence in my later testimony to the Senate on a later occasion (which he could not rebut or answer in any way), it appears that Mr Johnson may have confused electric Azipods, which are used on large ocean-going ships such as cruise ships, and the Navy's new LHDs, with pumpjets.  Azipods have open-tipped propellers, and are not remotely related to pumpjets.  So far as I can discern, this particular part of the testimony was made-up entirely impromptu without any knowledge or care for the truth of the facts.  The degree of confidence that we can have in his assurance that the pump-jet will be the optimum approach for the submarine should be considered in light of this.

\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{9 February 2018 - Department of Defence, in response to Question on Notice from Senator Rex Patrick in December 2017}]
\textit{Senator Patrick inquires about the nature, cost, and conclusions of earlier research into pumpjets undertaken as part of \$214 million worth of ``de-risking" studies already undertaken, as well as the schedule and cost of subsequent research and development of the pumpjet.  Prior research consisted of a single 99 page report obtained via literature review and discussion with experts at a cost of \$68,825.  Ongoing research included \$1.1million spent to date to develop a generic submarine and pumpjet design, compuational and analytic models, and validation of these models through experimental studies.  The development and test program is not scheduled to commence until the last quarter of 2018.  Full questions and responses can be found in \ref{QON148}, \ref{QON150}, \ref{QON151}}
\\
\\
``The study concluded that pumpjets have the potential to deliver increased propulsive efficiency and lower shaft torque requirements, relative to an open propeller. This should lead to beneficial platform systems and operational impacts to increase vessel speed and/or submerged endurance.
The study developed a parametric model of pumpjet thrust and efficiency that has been used as the basis for more recent studies into pumpjet performance."
\\
\\
``Specifications for the pump jet are being developed as design of the Future Submarine proceeds. Initial inputs will be developed in the last quarter of 2018, which marks the beginning of the development and test program for the pump jet. Development, testing, and assurance will be conducted by several parties including Naval Group, Direction générale de l'armement (DGA), and DSTG. Testing and assurance activities will take place in France (including at the premises of the DGA Techniques Hydrodynamiques in Val de Reuil) and Australia. Testing will be largely completed by CDR."
